Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-02-13 paper2.qmd
section:theoretical; process:write,figures

<!-- 4. what is the organization of the study? -->
As a result, the study divides its content into six main sections. @sec-thurstone_theory provides an overview of Thurstone's theory. @sec-theory-issues discusses the identified issues in detail. @sec-theoretical develops Thurstone's theory to address these challenges. The development integrates core theoretical principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. @sec-statistical translates these theoretical and practical elements into a probabilistic statistical model for the analysis of dichotomous pairwise comparison data. @sec-discussion discusses the implications of the findings and explores avenues for future research. Finally, @sec-conclusion summarizes the study's key insights.

@@

# Developing Thurstone's theory {#sec-theoretical}

@@

<!-- 3. Example of a CJ estimand and estimator, and its limitations -->
A motivating example helps to clarify these concepts. This example will appear throughout the document to illustrate various aspects of the proposed development. In this example, researchers aim to determine: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, researcher design a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method. Judges then compare pairs of students’ written texts (stimuli), to produce a dichotomous outcome that reflects the relative quality of each text (trait). Based on this setup, researchers reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?". Finally, following current CJ practices, researchers then rely on estimates from the BTL model, or its transformations, to approximate this estimand.

@@

<!-- 6. What the section will do? -->
Thus, this section addresses the issues identified in @sec-theory-issues by developing  Thurstone’s theory using the structural approach to causal inference. Specifically, it leverages the capabilities of this approach to formalize the combination of the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. In addition to enhancing statistical accuracy and strengthening measurement reliability and validity, the approach offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ experiments. Second, it shifts the current comparative data analysis paradigm from passively accepting the model assumptions to actively testing whether those assumptions fit the data under analysis.

@@

In line with Thurstone's theory, the other variables present in [SCM -@fig-cj03_scm] include the texts' discriminal processes $(T_{ha}, T_{ib})$ and their discriminal difference $(D_{hiabjk})$ (see @sec-thurstone_theory). Additionally, based on the arguments presented in @sec-theory-issue1b and the recommendations of @Andrich_1978 and @Wainer_et_al_1978, the [SCM -@fig-cj03_scm] incorporates the judges' biases $(B_{kj})$. Together with the outcome, these variables form the preliminary set of endogenous variables, $V = \{ O_{hiabjk}, D_{hiabjk}, T_{ha}, T_{ib}, B_{kj} \}$. Finally, the SCM shows the preliminary set of structural equations, $F = \{ f_{O}, f_{D} \}$, which define the non-parametric dependencies among these variables.

@@

Conceptual-population model, preliminary integration of theoretical principles.
:::

<!-- 4. The DAGs -->
Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020]. A DAG is a graph consisting of nodes connected by edges, where nodes represent random variables. The term *directed* indicates that edges or arrows extend from one node to another, showing the direction of causal influence. The term *acyclic* means that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020]. DAGs conventionally depict observed variables as solid black circles and unobserved (latent) variables as open circles [@Morgan_et_al_2014]. Moreover, they often omit exogenous variables for simplicity, following the *standard representation*. However, this section adopts the *magnified representation* of DAGS, as including exogenous variables can enhance clarity and reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020].

<!-- 5. the preliminary DAG for CJ -->
@fig-cj03_dag displays the DAG associated with [SCM -@fig-cj03_scm]. This graph illustrates the direction of causal influences among variables, reflecting the assumptions of Thurstone's theory and the practical aspects of CJ experiments. It also distinguishes between observable endogenous variables, such as the outcome, and variables considered latent within the CJ framework, such as the text's discriminal processes, their discriminal difference, and the judges' biases.

<!-- 6. Why is useful to declare the data structure? -->
Although specifying a system's data structure is not mandatory for using SCMs and DAGs, defining one can sometimes enhance clarity and facilitate the description of the system. Thus, to simplify the description of the CJ system, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as follows:

@@

Readers can better visualize the structure of these vectors and matrices by considering small values for the parameters corresponding to number of individuals $(n_{I} = 5)$, the number of available stimuli per individual $(n_{A} = 2)$, the total number of judges $(n_{J} = 3)$, and the number of judgments each judge performs $(n_{K} = 3)$. Using these values, the matrices described in equations ([-@eq-mat12]) and ([-@eq-mat3]) take the following form:

@@

<!-- 8. how de we use that data structure? -->
Thus, using the data structure defined by the matrices in equations ([-@eq-mat12]) and ([-@eq-mat3]), we can now re-express [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in an equivalent vectorized form, as shown in @fig-cj04. In this representation, the outcome $O^{cp}_{RPUV}$, the texts' discriminal difference $D_{RPUV}$, their discriminal processes $T_{IA}$, and the judges' biases $B_{JK}$ are no longer scalar values but vectors. These vectors capture all the values from the conceptual population. Specifically, $O^{cp}_{RPUV}$ and $D_{RPUV}$ are vectors of length ${n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$, while $T_{IA}$ and $B_{JK}$ are vectors of length $(n_{I} \cdot n_{A})$ and $(n_{J} \cdot n_{K})$, respectively.

@@

Conceptual-population model, vectorized form of [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag].
:::

<!-- 9. what else? -->
Building on the arguments developed in @sec-theory-issue2 and the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp. 138] and Item Response Theory (IRT) [@Fox_2010, chap. 6; @vanderLinden_et_al_2017_I, chap. 24], the conceptual-population model includes two hierarchical structural components. Each component defines how different observed or latent variables influence the primary latent variable of interest [@Everitt_et_al_2010]. This addition facilitates analyses that integrate the trait measurement and hypothesis testing processes, simultaneously accounting for the traits and their associated uncertainties, thereby reducing bias and enhancing the measurement precision.

@@

@fig-cj09 displays the two hierarchical structural components for the motivating example. Specifically, the top branch of [DAG -@fig-cj09_dag] depicts the first component. This component describes how the students' writing-quality trait $T_{I}$ and a set of *relevant* text-related variables $X_{IA}$ causally influence the texts' written-quality trait $T_{IA}$. The DAG also highlights the idiosyncratic errors $e_{IA}$, which capture variations in the texts' trait that neither the students' trait nor other relevant text-related variables can explain. Here, $T_{I}$ and $e_{IA}$ are latent vectors of length $n_{I}$ and $(n_{I} \cdot n_{A})$, respectively. Additionally, $X_{IA}$ is an observed matrix with a dimension of $(n_{I} \cdot n_{A})$ rows and $p_{IA}$ columns.

Furthermore, the top branch of [DAG -@fig-cj09_dag] describes how a set of *relevant* student-related variables $X_{I}$ causally influence the students' writing-quality trait $T_{I}$. The DAG also includes the idiosyncratic errors $e_{I}$, which capture variations in the the students' traits that the relevant student-related variables cannot explain. Here, $e_{I}$ is a latent vector of length $n_{I}$, and $X_{I}$ is an observed matrix with $n_{I}$ rows and $p_{I}$ columns.

@@

Conceptual-population model, full integration of theoretical principles.
:::

Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component. This component describes how the judges' bias $B_{J}$ and a set of *relevant* judgment-related variables $Z_{JK}$ causally influence the biases associated with each text $B_{JK}$. The DAG also highlights the idiosyncratic errors $e_{JK}$, which capture variations in the texts' bias that neither the judges' bias nor other relevant judgment-related variables can explain. In this description, $B_{J}$ and $e_{JK}$ are latent vectors of length $n_{J}$ and $(n_{J} \cdot n_{K})$, respectively. Additionally, $Z_{JK}$ is an observed matrix with a dimension of $(n_{J} \cdot n_{K})$ rows and $p_{JK}$ columns.

Moreover, the bottom branch of [DAG -@fig-cj09_dag] describes how a set of *relevant* judge-related variables $Z_{J}$ causally influence the judges' bias $B_{J}$. The DAG also accounts for idiosyncratic errors $e_{J}$, capturing the variations in the judges' bias that relevant judge-related variables cannot explain. Here, $e_{J}$ is a latent vector of length $n_{J}$, and $Z_{J}$ is an observed matrix with $n_{J}$ rows and $p_{J}$ columns. Notably, all variables and functions shown in [SCM -@fig-cj09_scm] and [DAG -@fig-cj09_dag] constitute the set of endogenous variables $V$ and structural equations $F$ for the conceptual-population model.

In summary,

@@

A cautious reader will notice that the model descriptions of @sec-theory-theoretical_P rely on several relevant variables. In this study, a *relevant* or a *sufficient adjustment* set of variables (potentially an empty set) refers to those that satisfy the *backdoor criterion* [@Neal_2020, pp. 37], that is, variables that block all non-causal paths without introducing new ones [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].

Incorporating these variables into the conceptual-population model enables researchers to identify different estimands of interest and their causal effects, even with observational data. This identification works because conditioning on these variables ensures that the estimator meets several critical properties, including confounding elimination [@Morgan_et_al_2014]. Here, *conditioning* means restricting the statistical analysis to a subset of the population defined by the values of a given variable [@Neal_2020, pp. 32].

Naturally, the validity of causal claims now hinges on the assumption that these variables serve as a sufficient adjustment set, a property that cannot be guaranteed for any variable. Nevertheless, as @Kohler_et_al_2019 [pp. 150] noted, drawing conclusions about the real world from observed data inevitably requires assumptions, and these requirements hold true for both observational and experimental data.

>>>>

<!-- 4. what is the organization of the study? -->
As a result, the study divides its content into six main sections. @sec-thurstone_theory provides an overview of Thurstone's theory. @sec-theory-issues discusses the identified issues in detail. @sec-theoretical extends Thurstone's theory to address these challenges. The extension integrates core theoretical principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. @sec-statistical translates these theoretical and practical elements into a probabilistic statistical model for the analysis of dichotomous pairwise comparison data. @sec-discussion discusses the implications of the findings and explores avenues for future research. Finally, @sec-conclusion summarizes the study's key insights.

@@

# Extending Thurstone's theory {#sec-theoretical}

@@

<!-- 3. Example of a CJ estimand and estimator, and its limitations -->
A motivating example helps to clarify these concepts. This example will appear throughout the document to illustrate various aspects of the proposed extension. In this example, researchers aim to determine: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, researcher design a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method. Judges then compare pairs of students’ written texts (stimuli), to produce a dichotomous outcome that reflects the relative quality of each text (trait). Based on this setup, researchers reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?". Finally, following current CJ practices, researchers then rely on estimates from the BTL model, or its transformations, to approximate this estimand.

@@

<!-- 6. What the section will do? -->
Thus, this section addresses the issues identified in @sec-theory-issues by extending  Thurstone’s theory using the structural approach to causal inference. Specifically, it leverages the capabilities of this approach to formalize the combination of the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. In addition to enhancing statistical accuracy and strengthening measurement reliability and validity, the approach offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ experiments. Second, it shifts the current comparative data analysis paradigm from passively accepting the model assumptions to actively testing whether those assumptions fit the data under analysis.

@@

In line with Thurstone's theory, [SCM -@fig-cj03_scm] depicts the texts' discriminal processes $(T_{ha}, T_{ib})$ and their discriminal difference $(D_{hiabjk})$ (see @sec-thurstone_theory). Additionally, based on the arguments in @sec-theory-issue1b and the recommendations of @Andrich_1978 and @Wainer_et_al_1978, the SCM incorporates the judges' biases $(B_{kj})$. Together with the outcome, these variables form the preliminary set of endogenous variables, $V = \{ O_{hiabjk}, D_{hiabjk}, T_{ha}, T_{ib}, B_{kj} \}$. Finally, the SCM shows the preliminary set of structural equations, $F = \{ f_{O}, f_{D} \}$, which define the non-parametric dependencies among these variables.

@@

Conceptual-population model, scalar form.
:::

<!-- 4. The DAGs -->
Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020]. A DAG is a graph consisting of nodes connected by edges, where nodes represent random variables. The term *directed* indicates that edges or arrows extend from one node to another, showing the direction of causal influence. The term *acyclic* means that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020]. DAGs conventionally depict observed variables as solid black circles and unobserved (latent) variables as open circles [@Morgan_et_al_2014]. Moreover, they often omit exogenous variables for simplicity, following a *standard representation*. However, this section adopts the *magnified representation* of DAGs, as including exogenous variables in their representation can enhance clarity and reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020].

<!-- 5. the preliminary DAG for CJ -->
@fig-cj03_dag displays the DAG corresponding to [SCM -@fig-cj03_scm], illustrating the causal relationships among variables based on Thurstone's theory and the practical aspects of CJ experiments. Particularly, the graph shows how the texts' discriminal processes $(T_{ha}, T_{ib})$ influence their discriminal difference $(D_{hiabjk})$, which in turn determines the outcome. It also depicts the judges' biases $(B_{kj})$ as an additional factor influencing the texts' discriminal difference. Furthermore, the DAG helps distinguish between observed endogenous variables, such as the outcome, and latent endogenous variables, which include the texts' discriminal processes, their discriminal difference, and the judges' biases.

<!-- 6. Why is useful to declare the data structure? -->
Although specifying a system's data structure is not mandatory for using SCMs and DAGs, defining one can sometimes enhance clarity and facilitate the description of the system. Thus, to simplify the description of a CJ experiment, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as follows:

@@

Readers can better visualize the structure of these vectors and matrices by considering small values for the parameters corresponding to number of individuals, the number of available stimuli per individual, the total number of judges, and the number of judgments each judge performs. For instance, assuming $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, the matrices described in equations ([-@eq-mat12]) and ([-@eq-mat3]) take the following form:

@@

<!-- 8. how de we use that data structure? -->
Thus, using the data structure defined by the matrices in equations ([-@eq-mat12]) and ([-@eq-mat3]), we can re-express [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in an equivalent vectorized form, as shown in @fig-cj04. In this depiction, the outcome $O^{cp}_{RPUV}$, the texts' discriminal difference $D_{RPUV}$, their discriminal processes $T_{IA}$, and the judges' biases $B_{JK}$ are represented as vectors rather than scalar values. These vectors encompass all values from the conceptual population. Specifically, $O^{cp}_{RPUV}$ is an observed vector of length ${n{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$, while $D_{RPUV}$ is a latent vector of the same dimension. Additionally, $T_{IA}$ and $B_{JK}$ are latent vectors of lengths $(n_{I} \cdot n_{A})$ and $(n_{J} \cdot n_{K})$, respectively.

@@

Conceptual-population model, vectorized form.
:::

<!-- 9. what else? -->
Moreover, building on the arguments developed in @sec-theory-issue2 and the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp. 138] and Item Response Theory (IRT) [@Fox_2010, chap. 6; @vanderLinden_et_al_2017_I, chap. 24], the conceptual-population model includes two hierarchical structural components. Each component defines how different observed or latent variables influence the primary latent variable of interest [@Everitt_et_al_2010]. This addition facilitates analyses integrating the trait measurement and hypothesis testing processes, simultaneously accounting for the traits and their associated uncertainties, thereby reducing bias and enhancing measurement precision.

@@

@fig-cj09 displays the two hierarchical structural components for the motivating example. Specifically, the top branch of [DAG -@fig-cj09_dag] depicts the first component. This component describes how the students' writing-quality trait $(T_{I})$ and a set of *relevant* text-related variables $(X_{IA})$ causally influence the texts' written-quality trait $(T_{IA})$. The DAG also highlights the idiosyncratic errors $(e_{IA})$, which capture variations in the texts' trait that neither the students' trait nor other relevant text-related variables can explain. Here, $T_{I}$ and $e_{IA}$ are latent vectors of length $n_{I}$ and $(n_{I} \cdot n_{A})$, respectively. Additionally, $X_{IA}$ is an observed matrix with a dimension of $(n_{I} \cdot n_{A})$ rows and $q_{IA}$ columns.

Furthermore, the top branch of [DAG -@fig-cj09_dag] describes how a set of *relevant* student-related variables $(X_{I})$ causally influence the students' writing-quality trait $(T_{I})$. The DAG also includes the idiosyncratic errors $(e_{I})$, which capture variations in the students' traits that the relevant student-related variables cannot explain. Here, $e_{I}$ is a latent vector of length $n_{I}$, and $X_{I}$ is an observed matrix with $n_{I}$ rows and $q_{I}$ columns.

@@

Conceptual-population model, final vectorized form.
:::

Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component. This component describes how the judges' bias $(B_{J})$ and a set of *relevant* judgment-related variables $(Z_{JK})$ causally influence the biases associated with each text $B_{JK}$. The DAG also highlights the idiosyncratic errors $(e_{JK})$, which capture variations in the texts' bias that neither the judges' bias nor other relevant judgment-related variables can explain. In this description, $B_{J}$ and $e_{JK}$ are latent vectors of length $n_{J}$ and $(n_{J} \cdot n_{K})$, respectively. Additionally, $Z_{JK}$ is an observed matrix with a dimension of $(n_{J} \cdot n_{K})$ rows and $q_{JK}$ columns.

Moreover, the bottom branch of [DAG -@fig-cj09_dag] describes how a set of *relevant* judge-related variables $(Z_{J})$ causally influence the judges' bias $(B_{J})$. The DAG also accounts for idiosyncratic errors $(e_{J})$, capturing the variations in the judges' bias that relevant judge-related variables cannot explain. Here, $e_{J}$ is a latent vector of length $n_{J}$, and $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ columns. Notably, all variables and functions shown in [SCM -@fig-cj09_scm] and [DAG -@fig-cj09_dag] constitute the set of endogenous variables $V$ and structural equations $F$ for the conceptual-population model.

Overall, the conceptual-population model extends Thurstone's theory by introducing key innovations to address the limitations discussed in @sec-theory-issues. These enhancements include accounting for judges' biases and integrating hierarchical structural components. Nevertheless, despite its statistical and measurement advantages, the model still relies on the unrealistic assumption that researchers can access data capturing all repeated judgments from every judge for each pair of stimuli produced by every individual in the population. Since this assumption is rarely met in practice, a more realistic scenario must be considered.

@@

<!-- A cautious reader will notice that the model descriptions of @sec-theory-theoretical_P rely on several relevant variables. In this study, a *relevant* or a *sufficient adjustment* set of variables (potentially an empty set) refers to those that satisfy the *backdoor criterion* [@Neal_2020, pp. 37], that is, variables that block all non-causal paths without introducing new ones [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].  -->

<!-- Incorporating these variables into the conceptual-population model enables researchers to identify different estimands of interest and their causal effects, even with observational data. This identification works because conditioning on these variables ensures that the estimator meets several critical properties, including confounding elimination [@Morgan_et_al_2014]. Here, *conditioning* means restricting the statistical analysis to a subset of the population defined by the values of a given variable [@Neal_2020, pp. 32]. -->

<!-- Naturally, the validity of causal claims now hinges on the assumption that these variables serve as a sufficient adjustment set, a property that cannot be guaranteed for any variable. Nevertheless, as @Kohler_et_al_2019 [pp. 150] noted, drawing conclusions about the real world from observed data inevitably requires assumptions, and these requirements hold true for both observational and experimental data. -->
